Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
section 2.3 edits

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.3em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent in the source language \eg{}
    \texttt{do} in Fortran for our implementation.} incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop. Induction variables are ranged
over by $i$, $j$, $k$ throughout. Since we do not use non-base
(derived) induction variables we often simply say \emph{induction variable}.
\end{definition}

@@

The \term{forward},
\term{backward}, and \term{centered} regions can all receive an
additional optional attribute \term{irreflexive} which marks
that the region does not include the origin (offset 0).
For example, the following is a \term{backward} stencil, which is similar
to \term{forward} but has negative neighbourhood indices, and which
is irreflexive:
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, irreflexive) :: a
b(i) = a(i-1) + a(i-2)
\end{minted}
%
A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{minted}

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \\
\multicolumn{2}{l}{\qquad\qquad\;\;\; \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}}  \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{minted}{fortran}
!= region r1 = forward(depth=1, dim=1)
!= region r2 = forward(depth=1, dim=2)
!= region robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is especially useful for common stencil patterns, such as Roberts cross,
as the region can be defined once and reused.
%%
\paragraph{Spatial modifiers}
%%
There are two modifiers: one used at specification and the other at region
level to provide better descriptions for stencil computations.

At the specification level, the \texttt{readOnce} modifier enforces that no
index appears more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
However, this would be an invalid specification if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion corresponds to that of \emph{linearity} in type systems,
though we opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

At the region level, \texttt{irreflexive} modifier qualifies base
regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
It indicates that in the dimension of the region there is no neighbour index
with $0$ offset, \eg{},
%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1, irreflexive) :: a
b(i) = a(i+1) + a(i-1)
\end{minted}
%
or with just a \texttt{forward} region an example would be:
%
\begin{minted}{fortran}
!= stencil forward(depth=1, dim=1, irreflexive) :: a
b(i) = a(i+1)
\end{minted}

Naturally, \term{irreflexive} qualifier cannot be used on \term{reflexive}
regions.

\paragraph{Under and over-approximation}

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil, which does not fit the rest
of the specification syntax. For example:
%%
\begin{minted}{fortran}
!= stencil atLeast, reflexive(dim=1)      :: a
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2)
\end{minted}
%%
%\subsubsection{Temporal specifications}

@@

The names ``forward'', ``backward'' and ``centered''
are inspired by standard terminology in numerical analysis
for the shape of discretisation schemes. For example,
the standard ``explicit method'' for approximation
PDEs is the \emph{Forward Time, Centered Space} (FTCS)
scheme~\cite{dawson1991finite}. For
the one-dimensional heat equations, an FTCS discretisation
provides approximation code with the following stencil~\cite{recktenwald2004finite}:
%%
\begin{minted}{fortran}
do i=2, n-1
  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
end do
\end{minted}
%%
where \texttt{r} and \texttt{r2} are some constants.
A valid specification for this stencil is then:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: v
\end{minted}
%%
Such a specification can be inferred, or can be inserted by a user
and checked against the code.

>>>>

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.3em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\vspace{-0.8em}
\end{figure}

@@

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop. Induction variables are ranged
over by $i$, $j$, $k$ throughout. Since we do not use non-base
(derived) induction variables we often simply say \emph{induction variable}.
\end{definition}

@@

The \term{forward},
\term{backward}, and \term{centered} regions can all receive an
additional optional attribute \term{irreflexive} which marks
that the region does not include the origin (offset 0).
For example, the following is a \term{backward} stencil, which is similar
to \term{forward} but has negative neighbourhood indices, and which
is irreflexive:
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, irreflexive) :: a
b(i) = a(i-1) + a(i-2)
\end{minted}
%
The \term{irreflexive} attribute cannot be used on \term{reflexive}
regions.

A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{minted}

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \\
\multicolumn{2}{l}{\qquad\qquad\;\;\; \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}}  \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\vspace{-0.5em}
\end{figure}

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{minted}{fortran}
!= region r1 = forward(depth=1, dim=1)
!= region r2 = forward(depth=1, dim=2)
!= region robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is especially useful for common stencil patterns, such as Roberts cross,
as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified 
by \emph{approximation} and \emph{multiplicity} information 
(in the \textit{spec} rule of Fig.~\ref{fig:syntax}).

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, in all of
the previous examples the \texttt{readOnce} modifier could be added,
\eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of \emph{linearity} in linear type
systems~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

%At the region level, \texttt{irreflexive} modifier qualifies base
%regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
%It indicates that in the dimension of the region there is no neighbour index
%with $0$ offset, \eg{},
%
%\begin{minted}{fortran}
%!= stencil centered(depth=1, dim=1, irreflexive) :: a
%b(i) = a(i+1) + a(i-1)
%\end{minted}
%
%or with just a \texttt{forward} region an example would be:
%%
%\begin{minted}{fortran}
%!= stencil forward(depth=1, dim=1, irreflexive) :: a
%b(i) = a(i+1)
%\end{minted}

%\paragraph{Under and over-approximation}

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil pattern, which cannot be expressed
precisely in our specification syntax. For example:
%%
\begin{minted}{fortran}
!= stencil atLeast, reflexive(dim=1)       :: a
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2)
\end{minted}
%%
%\subsubsection{Temporal specifications}

@@

The names ``forward'', ``backward'' and ``centered''
are inspired by standard terminology in numerical analysis
for the shape of discretisation schemes. For example,
the standard \emph{explicit method} for approximating
PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
scheme~\cite{dawson1991finite}. For
the one-dimensional heat equation, an FTCS discretisation
provides approximation code with the following stencil~\cite{recktenwald2004finite}:
%%
\begin{minted}{fortran}
do i=2, n-1
  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
end do
\end{minted}
%%
where \texttt{r} and \texttt{r2} are constants.
A valid specification for this is:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: v
\end{minted}
%%
Such a specification can be inferred, or can be inserted by a user
and checked against the code.
